Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-08 article.Rmd
Deleted extract
minor editing

This paper addresses two trends in the humanities: the rise of gender as an analytical tool and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name from historical data in order to take into account how naming practices change over time and space. In contrast to existing methods of predicting gender, this method provides a higher level of accuracy and a precise statement of the likelihood that a name is a given gender. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). The article demonstrates the method's utility by applying it to a case study in which we study gender in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*.

@@

Many other names as used in the United States, such as Madison, Morgan, Sydney, and Kendall, have also changed genders. Such changes in the prevailing gender of a name usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely happens in the other direction. In general, names that move from male to female also change from being unpopular to popular names. For example the name Madison changed from male to female suddenly after the 1985 because the name became extremely popular for naming baby girls, whereas before it had been an infrequently used male name. In other words, since the late twentieth century US at any rate, girls are more likely than boys to be receive a trendy name, and once a trendy name is associated with females it is unlikely to be used for males. The few exceptions of names which go from being female to male, such as Jan, have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommong, other causes besides changing gender preferences can explain the change.

@@

----

Things cut to be moved to more suitable places:

[NOTE: The SSA data includes only names used more than 5 times in a given year, so the "long tail" of actual number of names in the US is considerably larger. Link to SSA data.]

>>>>

This paper addresses two trends in the humanities: the rise of gender as an analytical tool and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name from historical data in order to take into account how naming practices change over time and space. In contrast to existing methods of predicting gender, this method provides a higher level of accuracy and a precise statement of the likelihood that a name is a given gender. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to demonstrate the method's utility by applying it to a case study in which we study gender in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*.

@@

Many other American names, such as Madison, Morgan, Sydney, and Kendall, have also changed gender over the past century. Such changes usually follow set patterns. In general, names tend to shift from being predominantly male to being predominantly female; the shift rarely moves in the other direction. In general, names that transition from male to female also became increasingly popular over time. For example, the name "Madison" changed from male to female suddenly after 1985. Whereas before it had been an infrequently used male name, after 1985 it skyrocketed in popularity, and by the early 2000s it was one of the ten most frequent names for baby girls. Broadly speaking, since the late twentieth century Americans girls are more likely than American boys to receive a "trendy" name, and once that name is associated with females it is unlikely to be given to male babies. The few exceptions of names which go from being female to male (such as "Jan") have a different causal mechanism. It seems likely that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change. **this last bit about "Jan" might just muddle things - unless we have a clear explanation or hypothesis as to what does explain the change, not sure it's worth highlighting**

@@
